Sampling requirements to detect ecosystem change

### 2020-03-30 R/manuscript_draft.Rmd
made small edits to text in discussion section

Outline:

- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2))
- money implications (before or now)  
- how this fits in with other work: examples of situation it would be applied (Rosalie)  
- how this doesn't fit with other work (limitations) (Rosalie)
- future work  (?)

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
There is often a difference between the number of samples analyzed and the total number of samples collected, due to time or funding limitations. When the processing time is disproportionally higher than the collection time especially, a subsampling can be done posterior to sampling. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}).

We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.  (Fig. \ref{fig:simulation_min_time_plot}).

@@

Multiple subsampling strategies can be chosen (Fig. \ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig. \ref{fig:simulation_approaches_plot}c). Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint.

Applied to the real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig. \ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples. Same results for the major change, with only 12% of the samples analyzed. The method also work to detect further shifts (Fig. S4).
@Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core. Each sample took an average of 3 hours to process. We found that using an interval approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work. This correspond to several thousands of US dollars depending on labor costs.

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred. Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint. It matches very well the paleoecological approach [@Wingard2017]. However, both different types of data and different questions can be addressed with the same approach. Suppose instead that the goal was to detect a change in observed individuals with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

@@

Our approach is able to address other data sources and questions, but not all. When processing time and resources are less than the collection time, investigators will be likely to process every sample. If the studied phenomenon follow the rules of a time-series, and the question concerns a changepoint, it is worth noting that a regular sampling is desirable compared to a random one (Figs. \ref{fig:simulation_approaches_plot},S2,S3). More generally, some questions require a continuous time series, or at least a regular sampling interval. For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016]. However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver.

\rosalie{I don't understand what you meant here. I tried reformulating some of it, but we can talk about it more if needed}:  
\st{The specific number of samples and optimal sampling approach will largely be system specific. A similar system may be used to get some sense for the best sampling scheme, but it will depend on the system dynamics (e.g. interannual variation, trend effect size). The specific sampling scheme also depends on the question of interest. For instance, changepoint detection may require fewer samples than detecting a change in biodiversity indices (citations). Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. However, more samples could be processed later to address these additional questions.}

@@

Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d]. For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014). We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint. This is valuable given the high costs of many monitoring programs [@Caughlan2001a,Bennett2014a]. Monitoring programs have to be able to address our questions of interest with sufficient statistical power. The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007,@Bennett2018b]. Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future.  
We make the point that if a subsample of the samples must be used, the temptation to be unbiased by using a regular subsampling will lead to losing a lot of accuracy in the process. If the objective is clearly to detect changepoint, an iterative, partial approach, is desirable.

>>>>

<!--
Outline:

- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2))
- money implications (before or now)  
- how this fits in with other work: examples of situation it would be applied (Rosalie)  
- how this doesn't fit with other work (limitations) (Rosalie)
- future work  (?)
-->

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionally higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.  (Fig. \ref{fig:simulation_min_time_plot}).

@@

Multiple subsampling strategies can be chosen (Fig. \ref{fig:conceptual}), but only iterative sampling detected the real changepoint with a limited number of samples (Fig. \ref{fig:simulation_approaches_plot}c). Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint. Applied to the real case study, the iterative method allowed to find the main changepoint with 9 samples analyzed (Fig. \ref{fig:case_study_number_sample}), with a starting point of 5 randomly distributed samples. The method also work to detect further shifts (Fig. S4).

@Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core. Each sample took an average of 3 hours to process. We found that using an interval approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work. This correspond to several thousands of US dollars depending on labor costs.

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019d,@White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred. Our interval sampling approach involved taking a fixed number of regularly-spaced samples to start and then adding more samples strategically to detect a changepoint. \easton{I'm not sure what this means. What do you mean by it ``matches"?}It matches very well the paleoecological approach [@Wingard2017]. However, both different types of data and different questions can be addressed with the same approach. Suppose instead that the goal was to detect a change in observed individuals with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample. When sample analysis costs are high, then regular and interative sampling are better than random sampling (Figs. \ref{fig:simulation_approaches_plot},S2,S3). However, some questions require a continuous time series, or at least a regular sampling interval. For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016]. However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. However, more samples could be processed later to address these additional questions.

@@

Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019d]. For example, a continuous sampling allows the calculation of autocorrelation or variance, that can be valuable indicators of critical slowing down (e.g., Dakos and Bascompte, 2014). We show here that an informed subsampling can still allow detection of critical information, such as time series changepoint. This is valuable given the high costs of many monitoring programs [@Caughlan2001a,@Bennett2014a]. Monitoring programs have to be able to address our questions of interest with sufficient statistical power. The costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007,@Bennett2018b]. Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future. If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling. This can improve the accuracy of the results and reduce costs overall. If the objective is to detect a changepoint, an iterative approach is optimal.
